Simultaneous segmentation and distortion correction on diffusion MRI data using shape priors

### 2012-12-22 Publications/2013-IPMI/Latex/00-Main.tex
Included Alessandro's suggestions
And re-structured the text regarding the model 1

\title{Simultaneous segmentation and distortion correction on diffusion weighted MR using shape priors}
%%
\titlerunning{Simultaneous segmentation and distortion correction on DWI}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
%\author{Oscar Esteban\inst{1,2} \and
%        Alessandro Daducci \inst{2} \and
%		Meritxell Bach-Cuadra \inst{2,3} \and
%        Jean-Philippe Thiran \inst{2,3} \and
%        Andr\'es Santos \inst{1} \and
%        Dominique Zosso \inst{2,4}}
%
%\authorrunning{Oscar Esteban et al.} % abbreviated author list (for running head)

>>>>

\title{Simultaneous segmentation and distortion correction on diffusion MRI data using shape priors}
%%
\titlerunning{Simultaneous segmentation and distortion correction on dMRI}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
%\author{Oscar Esteban\inst{1,2} \and
%        Alessandro Daducci \inst{2} \and
%		Meritxell Bach-Cuadra \inst{2,3} \and
%        Jean-Philippe Thiran \inst{2,3} \and
%        Andr\'es Santos \inst{1} \and
%        Dominique Zosso \inst{2,4}}
%
%\authorrunning{Oscar Esteban et al.} % abbreviated author list (for running head)

### 2012-12-22 Publications/2013-IPMI/Latex/03-Introduction.tex
Included Alessandro's suggestions
And re-structured the text regarding the model 1

The whole-brain connectivity analysis has given rise to some challenges
towards reliable structural information
about the neuronal tracts from \gls{dwi} \cite{johansen-berg_using_2009,
jones_white_2012}. The earlier stages of these
processing pipelines generally include two necessary steps, brain tissue
segmentation on the diffusion space and the correction of geometrical
distortions inherent to the imaging techniques \citep{hagmann_mr_2012}.

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \gls{csf}-\gls{gm} and \gls{gm}-\gls{wm} interface surfaces.
This segmentation is an processing step on which strongly rely further
tasks. In tractography, a high-standard \gls{wm} mask is required. Otherwise,
there is an important risk for the algorithm to lose fiber bundles. This
requirement is usually solved in practice by plainly thresholding the 
\gls{fa}, a well-known scalar map derived from \gls{dwi} which depicts 
the isotropy of water diffusion inside the brain. 
A precise location of the
\gls{gm}-\gls{wm} surface is also essential in the final steps to
achieve a consistent parcellisation of the cortex to represent the nodes 
of the output network. This parcellisation is generally defined in a 
high-resolution and better understood structural \gls{mri} of the same 
subject (e.g. \gls{t1} and/or \gls{t2} weighted acquisitions). %Conversely, this problem is resolved with non-linear registration of a structural \gls{mri} of the subject to the \gls{dwi} data. 
Even though some efforts have addressed
the study of the robustness of tractography with respect to intra-subject variability
\cite{heiervang_between_2006,wakana_reproducibility_2007}, these results are restricted to some regions of the brain, only. Therefore, robust
and precise segmentation methods are required in the whole-brain application.

The \gls{dwi} data is usually obtained with \gls{epi}
acquisition techniques, that often suffer from severe distortions due to 
local field inhomogeneities. Generally, it is appreciated in the anterior
part of the brain, along the phase-encoded direction. Some methodologies have
been developed and generically named as \emph{\gls{epi}-unwarp} techniques
\cite{holland_efficient_2010,hsu_correction_2009,jezzard_characterization_2005,
reber_correction_2005}. These methods usually 
require the extra acquisition of the magnitude and phase of
the field (``field-mapping''), a condition which is not always met. Some other 
methodologies do not make use of the field-mapping, compensating the distortion
with non-linear registration from structural \gls{mri} or other means
\citep{andersson_modeling_2001}. To our knowledge, there exists no study
on the impact of the \gls{epi} distortion on the variability of tractography
results.

Therefore, the problems of precise segmentation in \gls{dwi}-space and the 
spatial mapping between these contours and the corresponding surfaces in 
anatomical images bear significant redundancy. Once the spatial relationship 
between \gls{t1} and \gls{dwi} space is established, the contours which are 
readily available in \gls{t1} space, can simply be projected on to the 
\gls{dwi}-data. Conversely, if a precise delineation in \gls{dwi}-space 
was achieved, the spatial mapping with \gls{t1}-space could be derived 
from one-to-one correspondences on the contours. However, neither segmentation 
nor registration can be performed flawlessly, if considered independently. 
The significant benefits of exploiting the anatomical \gls{mri} when 
segmenting the \gls{dwi} data have been demonstrated \cite{zollei_improved_2010}, 
justifying the use of the shape prior information.

We suggest clustering the current methodologies of template-based segmentation 
methods into three groups. The first group typically adds a shape prior term to 
the energy functional of an evolving active contour~\citep{Bresson2006a,Chan2005,
Chen2002,Cremers2006,Gastaud2004,Paragios2003,Vemuri2003a,Yezzi2003a} %removed Rousson2002,Ayvaci2007,Schmid2008
%{\color{red}{(FIXME:I would select from here the most important,
%removing directly those appearing in conferences and never published on journal. Better
%if we found a review paper/book chapter. I don't like so many references on a row)}}.
These methods generally have a explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. Closely related to this group are atlas-based segmentation
methods \citep{Gorthi2011,Gorthi2009,Pohl2005,Pohl2006,Wang2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions. Here, the 
presence of more structures than one unique \gls{roi} helps aligning the target image 
with the atlas in a hierarchical fashion. Finally, the third group generalizes 
the atlas to actual images, and the contour is to segment simultaneously two 
different target images, related by a spatial transform to be co-estimated~\citep{Wyatt2003,
Yezzi2003}.

In this paper we propose a novel registration framework to simultaneously
solving the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed anatomy extracted from anatomical \gls{mri}. Indeed,
hereafter we consider the segmentation in anatomical images as a solved problem.
Moreover, the shape prior is of very ``strong'' nature, since it is specific to 
the particular subject. Also, after global alignment using existing approaches, 
the remaining spatial deformation between anatomical and diffusion space is 
due to \gls{mr} distortion. Finally, we need to establish precise spatial 
correspondence between the surfaces in both spaces, including the tangential 
direction for parcellation. Therefore, we can reduce the problem to finding 
the differences of spatial distortion in between anatomical and \gls{dw} space.
We thus reformulate the segmentation problem as an inverse problem, where we 
seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space, such that the structural 
contours segment optimally the \gls{dwi} data. In the process, the one-to-one 
correspondence between the contours in both spaces is guaranteed, and projection 
of parcellisation to \gls{dw} space is implicit and consistent.

>>>>

The whole-brain connectivity analysis has given rise to some challenges
towards reliable structural information about the neuronal tracts 
from \gls{dwi} \cite{johansen-berg_using_2009,jones_white_2012}. 
The earlier stages of these processing pipelines generally include 
two necessary steps, brain tissue segmentation on the diffusion space 
and the correction of geometrical distortions inherent to the acquisition 
sequence \citep{hagmann_mr_2012}.

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \gls{csf}-\gls{gm} and \gls{gm}-\gls{wm} interface surfaces.
This segmentation is a processing step on which strongly rely further
tasks. An accurate brain tissue segmentation is required
to filter the fiber bundles obtained with \gls{dwi} tractography. 
This requirement is usually solved in practice by plainly thresholding the 
\gls{fa}, a well-known scalar map derived from \gls{dwi} which depicts 
the isotropy of water diffusion inside the brain. Also,
it is necessary to locate the intersections of fiber bundles and \gls{gm}.
Moreover, a precise location of the \gls{gm}-\gls{wm} surface is also essential 
to finally achieve a consistent parcellisation of the cortex 
to represent the nodes of the output network. This parcellisation 
is generally defined in a high-resolution and better understood structural 
\gls{mri} of the same 
subject (e.g. \gls{t1} and/or \gls{t2} weighted acquisitions). Even though
some efforts have addressed the study of the robustness of tractography with
respect to intra-subject variability \cite{heiervang_between_2006,
wakana_reproducibility_2007}, these results are restricted to certain regions 
of the brain, only. Therefore, robust and precise segmentation methods are 
required in the whole-brain application. The problem is challenging due to
the much lower resolution of \gls{dwi} (typically around 
$2.5\times2.5\times2.5mm^3.$) compared to structural \gls{mri}, and the
existence of geometrical distortions.

\gls{dwi} data are usually acquired with \gls{epi} sequences 
as they allow for very fast acquisitions but they are known to 
suffer from geometrical distortions due to local field inhomogeneities. 
These artifacts happen along the phase-encoding direction, and are most 
appreciable in the front part of the brain for the strong air/tissue 
interface around the frontal sinuses. Some methodologies have
been developed and generically named as \emph{\gls{epi}-unwarp} techniques
\cite{holland_efficient_2010,hsu_correction_2009,jezzard_characterization_2005,
reber_correction_2005}. These methods usually 
require the extra acquisition of the magnitude and phase of
the field (``field-mapping''), a condition which is not always met. Some other 
methodologies do not make use of the field-mapping, compensating the distortion
with non-linear registration from structural \gls{mri} or other means
\citep{andersson_modeling_2001}. To our knowledge, there exists no study
on the impact of the \gls{epi} distortion on the variability of tractography
results.

Therefore, the problems of precise segmentation in \gls{dwi}-space and the 
spatial mapping between these contours and the corresponding surfaces in 
anatomical images bear significant redundancy. Once the spatial relationship 
between \gls{t1} and \gls{dwi} space is established, the contours which are 
readily available in \gls{t1} space can simply be projected on to the 
\gls{dwi}-data. Conversely, if a precise delineation in \gls{dwi}-space 
was achieved, the spatial mapping with \gls{t1}-space could be derived 
from one-to-one correspondences on the contours. However, neither segmentation 
nor registration can be performed flawlessly, if considered independently. 
The significant benefits of exploiting the anatomical \gls{mri} when 
segmenting the \gls{dwi} data have been demonstrated \cite{zollei_improved_2010}, 
justifying the use of the shape prior information.

We suggest clustering the current methodologies of template-based segmentation 
methods into three groups. The first group typically adds a shape prior term to 
the energy functional of an evolving active contour~\citep{Bresson2006a,Chan2005,
Chen2002,Cremers2006,Gastaud2004,Paragios2003,Vemuri2003a,Yezzi2003a}.
These methods generally have a explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. Closely related to this group are atlas-based segmentation
methods \citep{Gorthi2011,Gorthi2009,Pohl2005,Pohl2006,Wang2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions. Here, the 
presence of more structures than one unique \gls{roi} helps aligning the target image 
with the atlas in a hierarchical fashion. Finally, the third group generalizes 
the atlas to actual images, and the contour is to segment simultaneously two 
different target images, related by a spatial transform to be co-estimated~\citep{Wyatt2003,
Yezzi2003}.

In this paper we propose a novel registration framework to simultaneously
solving the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed morphology extracted from high-resolution anatomical 
\gls{mri}. Indeed, hereafter we assume this segmentation problem in anatomical 
images as a solved by widely used available procedures.
Moreover, the shape prior is of very ``strong'' nature, since it is specific to 
the particular subject. Also, after global alignment using existing approaches, 
the remaining spatial deformation between anatomical and diffusion space is 
due to \gls{epi} distortions. Finally, we need to establish precise spatial 
correspondence between the surfaces in both spaces, including the tangential 
direction for parcellation. Therefore, we can reduce the problem to finding 
the differences of spatial distortion in between anatomical and \gls{dw} space.
We thus reformulate the segmentation problem as an inverse problem, where we 
seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space, such that the structural 
contours segment optimally the \gls{dwi} data. In the process, the one-to-one 
correspondence between the contours in both spaces is guaranteed, and projection 
of parcellisation to \gls{dw} space is implicit and consistent.

### 2012-12-22 Publications/2013-IPMI/Latex/045-Experiments.tex
Included Alessandro's suggestions
And re-structured the text regarding the model 1

%
\subsection{Simulated diffusion data}
%
In order to demonstrate the functionality of the methodology, 
and characterize its possibilities with diffusion data,
we simulated the \gls{dw} signal of a synthetic phantom from a model
consisting of several spherical shapes emulating
the different brain tissues (see \autoref{fig:fa}, first row). 
We reconstructed the signal with standard procedures to 
approximate the environment to the real one at maximum. \\

\paragraph{Signal simulation}
To numerically simulate the MRI signal attenuation when applying a diffusion 
gradient in a voxel with $N$ fiber populations we made use of the standard 
\emph{Multi-Tensor Model}~\cite{Tuch:2002aa}:
%
\begin{equation} 
\label{eqn:MultiTensor}
S(\vec{q}) / S_{0} = \sum_{i=1}^{N} f_{i} \, \exp{ \left( -b \, \vec{q}^{T} \, \mathbf{D}_i \, \vec{q}\right) } + f_{iso} \, \exp{ \left( -b \, \mathbf{D}_{iso} \right) } ,
\end{equation}
%
where $\vec{q} \in \mathbb{S}^2$ is the direction of the diffusion gradient 
applied, $b$ is the b-value accounting for its strength, $S_0 \equiv S(\vec{0})$ 
is the signal with no diffusion weighting, $f_i$ and $\mathbf{D}_i$ are 
respectively the volume fraction and the diffusion tensor characterizing the 
$i$-th fiber population.

The diffusion tensors $\mathbf{D}_i$ and $\mathbf{D}_{iso}$ describe the diffusion 
processes of each fiber compartment and of contaminations from the CSF. In this work, 
these quantities have been taken from standard ranges typically observed in the 
human brain~\cite{Canales-Rodriguez:2009aa}.

@@

\paragraph{Derived scalar features}
The target \gls{dwi} data is characterized by its distortions and its
low resolution (typically around $2.2x2.2x3mm^3$). Depending on the
posterior reconstruction methodology and the angular resolution
intended, the \gls{dwi} raw data has to be processed in order to
extract the information in a manageable manner. The properties of
the reconstructed tensors and derived scalar maps have been
studied by \cite{ennis_orthogonal_2006}. Based on their
findings, the proposed energy model adapts to the \gls{fa}~\eqref{eq:fa} 
and \gls{md}~\eqref{eq:md} for their properties. 
Whereas \gls{fa} \ describes the \emph{shape} of diffusion, 
the \gls{md} depicts the \emph{magnitude} of the process.

\begin{align}
\mathrm{FA} &= \sqrt{ \frac{1}{2}}\,\frac{\sqrt{ (\lambda_1 - \lambda_2)^2 + (\lambda_2 - \lambda_3)^2 + (\lambda_3 - \lambda_1)^2}}{\sqrt{ {\lambda_1}^2 + {\lambda_2}^2 + {\lambda_3}^2}} \label{eq:fa} \\
\mathrm{MD} &= ( \lambda_1 + \lambda_2 + \lambda_3 ) / 3 \label{eq:md}
\end{align}
where $\lambda_i$ are the eigenvalues of the diffusion tensor 
associated with the diffusion signal $S(\vec{q})$. There exist 
two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \gls{dwi}-derived scalar maps. \\

\subsubsection{Simulated distortions}

For this model, we created manually a sound distortion visually similar
to real \gls{epi} distortions. We interpolated the distortion to a 
dense deformation field, necessary for warping the raw \gls{dwi} simulated
data. Once the signal was deformed, we proceeded to reconstruct the
\gls{dti} and subsequently obtained the scalars of interest (\gls{fa}, \gls{md}).
Finally, we estimated their parameters using the tissue probability
distribution maps from the original model (\autoref{table:parameters}).

>>>>

%
\subsection{Simulated diffusion data}
%
In order to demonstrate the functionality of the methodology, 
and characterize its possibilities with diffusion data,
we built a synthetic phantom from a model consisting of several 
spherical shapes emulating the different brain tissues (see 
\autoref{fig:fa}, first row). We simulated a single-shell
acquisition at $b=1000$ \eqref{eqn:MultiTensor} with 30 samples
equally distributed on one hemisphere, corresponding to a standard
\gls{dti} acquisition commonly used in clinical practice.
We generated a synthetic displacement field to produce
an \gls{epi}-like distortion on the \gls{dwi} signal. Finally,
we reconstructed the deformed signal, and generated \gls{dwi}-derived
scalar maps (specifically the \gls{fa} and \gls{md} maps).
Then, the scalar maps were placed as features in 
\eqref{eq:complete_energy} to drive our segmentation approach, 
finding the location of the \gls{wm}-\gls{gm}
and the \gls{csf}-\gls{wm} interfaces in the distorted space. \\

\paragraph{Signal simulation}
To numerically simulate the \gls{mri} signal attenuation when applying a diffusion 
gradient in a voxel with $N$ fiber populations we made use of the standard 
\emph{Multi-Tensor Model}~\cite{Tuch:2002aa}:
%
\begin{equation} 
\label{eqn:MultiTensor}
S(q) / S_{0} = \sum_{i=1}^{N} f_{i} \, \exp{ \left( -b \, q^{T} \, \mathbf{D}_i \, q\right) } + f_{iso} \, \exp{ \left( -b \, \mathbf{D}_{iso} \right) }  + f_{gm} \, \exp{ \left( -b \, \mathbf{D}_{gm} \right) } ,
\end{equation}
%
where $q \in \mathbb{S}^2$ is the direction of the diffusion gradient applied,
$b$ is the b-value accounting for its strength and $S_0 \equiv S(0)$ is the 
signal with no diffusion weighting. $f_i$ and $\mathbf{D}_i$ are the volume 
fraction and the diffusion tensor characterizing the $i$-th fiber population, 
whereas the tensors $\mathbf{D}_{iso}$ and $\mathbf{D}_{gm}$ describe the 
diffusion processes of partial volumes with \gls{csf} and \gls{gm} within the 
voxel, which volume fractions are, respectively, $f_{iso}$ and $f_{gm}$, and
$\sum_i f_i + f_{iso} + f_{gm} = 1$. In this work, the diffusion properties have
been taken from standard ranges typically observed in the human 
brain~\cite{Canales-Rodriguez:2009aa}. Moreover we assumed $N = 1$ and $S_0 = 1$ 
without loss of generality.

@@

\paragraph{Simulated \gls{epi} distortion}
For this model, we created manually a sound distortion to approximate
the real \gls{epi} distortions. We interpolated the distortion to a 
dense deformation field, and applied it to generate the deformed \gls{dwi}
signal.

\paragraph{Derived scalar features}
We obtained the local fiber configuration in each voxel with a commonly 
used \gls{dti} reconstruction tool~\footnote{DTIFIT, included
in the FMRIB's Software Library (FSL), 
\url{http://fsl.fmrib.ox.ac.uk/fsl/fsl-4.1.9/fdt/fdt_dtifit.html}}.

The properties of
the reconstructed tensors and derived scalar maps have been
studied by \cite{ennis_orthogonal_2006}. Based on their
findings, we chose \gls{fa}~\eqref{eq:fa} 
and \gls{md}~\eqref{eq:md} as features for the energy model in driving
the registration-segmentation process. 
Whereas \gls{fa} \ describes the \emph{shape} of diffusion, 
the \gls{md} depicts the \emph{magnitude} of the process.

\begin{align}
\mathrm{FA} &= \sqrt{ \frac{1}{2}}\,\frac{\sqrt{ (\lambda_1 - \lambda_2)^2 + (\lambda_2 - \lambda_3)^2 + (\lambda_3 - \lambda_1)^2}}{\sqrt{ {\lambda_1}^2 + {\lambda_2}^2 + {\lambda_3}^2}} \label{eq:fa} \\
\mathrm{MD} &= ( \lambda_1 + \lambda_2 + \lambda_3 ) / 3 \label{eq:md}
\end{align}
where $\lambda_i$ are the eigenvalues of the diffusion tensor 
associated with the diffusion signal $S(\vec{q})$. There exist 
two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \gls{dwi}-derived scalar maps. The properties
of brain tissues on the joint distribution of \gls{fa} and \gls{md} are 
summarized in \autoref{table:parameters}. \\

### 2012-12-22 Publications/2013-IPMI/Latex/05-Results.tex
Included Alessandro's suggestions
And re-structured the text regarding the model 1

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{model2result_b_1} &
\includegraphics[width=0.19\textwidth]{model2result_b_2} &
\includegraphics[width=0.19\textwidth]{model2result_a_1} &
\includegraphics[width=0.19\textwidth]{model2result_a_2} &
\includegraphics[width=0.19\textwidth]{model2surf} \\
\multicolumn{2}{c}{a)} & \multicolumn{2}{c}{b)} & c)
\end{tabular}
\caption{Result of the segmentation performance on the sulcus model.
a) Two slices and prior contours before segmentation-registration. b) The same slices and contours after distortion estimation. c) 3D rendering of the deformed priors, with the traces of the two selected slices highlighted. 
%In red and blue colors the obtained \gls{wm}/\gls{gm} 
%and \gls{gm}/\gls{csf} contours for the deformed image, respectively.
%In white color, the initial shape priors for the \gls{wm}/\gls{gm}
%interface.
{\color{red}{This figure is to be replaced by the real result,
this is a proof of concept of registration that does not solve
the pv problem. The current figure also removes the outer prior}}}
\label{fig:sulcusmodel_result}
\end{figure}

>>>>

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{model2result_b_1} &
\includegraphics[width=0.19\textwidth]{model2result_b_2} &
\includegraphics[width=0.19\textwidth]{model2result_a_1} &
\includegraphics[width=0.19\textwidth]{model2result_a_2} &
\includegraphics[width=0.19\textwidth]{model2surf} \\
\multicolumn{2}{c}{a)} & \multicolumn{2}{c}{b)} & c)
\end{tabular}
\caption{Result of the segmentation performance on the sulcus model.
a) Two slices and prior contours before segmentation-registration. b) The same slices and contours after distortion estimation. c) 3D rendering of the deformed priors, with the traces of the two selected slices highlighted.}
\label{fig:sulcusmodel_result}
\end{figure}

### 2012-12-22 Publications/2013-IPMI/Latex/06-Conclusion.tex
Included Alessandro's suggestions
And re-structured the text regarding the model 1

We implemented the methodology upon the widely used
Insight Registration and Segmentation 
Toolkit~\footnote{\url{http://www.itk.org}} (ITK)
for its computational benefits, the standardized code, and 
with the aim at making the procedure publicly available 
when ready for sharing with the research community.

>>>>

We implemented the methodology upon the widely used
Insight Registration and Segmentation 
Toolkit~\footnote{\url{http://www.itk.org}} (ITK)
for its computational benefits, the standardized code, and 
with the aim at making the procedure publicly available 
when ready for sharing with the research community.\\
